qcircuit 2.5.3 Tutorial

### 2017-09-25 CO681-QuantumInformationProcessing/assignments/assignment_01.tex
solved 4

@@

>>>>

\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
\DeclarePairedDelimiter\norm{\lVert}{\rVert}%

@@

Suppose that a qubit is prepared in the state $\ket{0}$ with probability $p$ and prepared in state $\ket{+}$ with probability $q$, where $p+q=1$ and $p,q\geq 0$. Then a measurement in the basis $\ket{\phi_0}, \ket{\phi_1}$ identifies the state successfully with probability
$$p\abs{\ip{\phi_0}{0}}^2 + q\abs{\ip{\phi_1}{+}}^2.$$
Since $\ket{0}$ and $\ket{+}$ lie in $\R^2$, the optimal $\ket{\phi_0}$ and $\ket{\phi_1}$ lie in $\R^2$ so as to optimize the length of projection of $\ket{\phi_0}$ and $\ket{\phi_1}$ onto $\ket{0}$ and $\ket{+}$ respectively (ie $\ip{\phi_0}{0}$ and $\ip{\phi_1}{+}$).
We can describe an arbitrary orthonormal basis of $\R^2$ as
$$\ket{\phi_0} = \begin{bmatrix} \cos\theta \\ \sin\theta \end{bmatrix}, \quad \ket{\phi_1}=\begin{bmatrix}-\sin\theta \\ \cos\theta \end{bmatrix}.$$
Now our success probability simplifies as
$$f(\theta) :=p\cos^2\theta + \frac{q}{2}(\cos\theta - \sin\theta)^2$$
with first derivative
$$f'(\theta) = -2p\sin\theta\cos\theta - q(\cos^2\theta - \sin^2\theta).$$
If we set $p=q=\frac{1}{2}$ then we have
$$f(\theta) = \frac{1}{2}\cos^2\theta + \frac{1}{4}(\cos\theta - \sin\theta)^2$$
and
$$f'(\theta) = -\sin\theta\cos\theta -\frac{1}{2}(\cos^2\theta-\sin^2\theta).$$
So $f$ is periodic with period $\pi$. Setting $f'(\theta)=0$ and solving we obtain one possible optimal solution: $\theta^* = \frac{-\pi}{8}.$
Since $f'(\frac{-2\pi}{8}) >0$ and $f'(0) < 0$, we know $\theta^*$ is a maximizer of $f$. Hence we have optimal success probability
$$f(\frac{-\pi}{8}) = \frac{1}{4}(2 + \sqrt(2)) \approx 0.85$$
and optimal basis
$$\ket{\phi_0} = \begin{bmatrix} \cos(\frac{-\pi}{8}) \\ \sin(\frac{-\pi}{8}) \end{bmatrix}, \quad \ket{\phi_1}=\begin{bmatrix}-\sin(\frac{-\pi}{8}) \\ \cos(\frac{-\pi}{8}) \end{bmatrix}.$$
If we set $p=\frac{3}{4}$ and $q=\frac{1}{4}$ then we have
$$f(\theta) = \frac{3}{4}\cos^2\theta + \frac{1}{8}(\cos\theta - \sin\theta)^2$$
and
$$f'(\theta) = -\frac{3}{2}\sin\theta\cos\theta -\frac{1}{4}(\cos^2\theta-\sin^2\theta).$$
So $f$ is periodic with period $\pi$. Setting $f'(\theta)=0$ and solving we obtain one possible optimal solution: $\theta^* = \arctan(3-\sqrt{10}).$
Since $f'(\arctan(3-\sqrt{10})-\frac{\pi}{4}) >0$ and $f'(\arctan(3-\sqrt{10}) + \frac{\pi}{4}) < 0$, we know $\theta^*$ is a maximizer of $f$. Hence we have optimal success probability
$$f(\arctan(3-\sqrt{10})) \approx 0.895$$
and optimal basis
$$\ket{\phi_0} = \begin{bmatrix} \cos(\arctan(3-\sqrt{10})) \\ \sin(arctan(3-\sqrt{10})) \end{bmatrix}, \quad \ket{\phi_1}=\begin{bmatrix}-\sin(\arctan(3-\sqrt{10})) \\ \cos(\arctan(3-\sqrt{10})) \end{bmatrix}.$$
